---
title: "FOP Analysis Rerun"
author: "Authors: Ethan Dinh"
date: "2025-10-2"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: '5'
    df_print: paged
  html_notebook:
    theme: spacelab
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes 
      smooth_scroll: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
---

```{css, echo=FALSE}
img {
  align-self: center;
  height: fit-content;
}

pre {
 margin: 20px 0;
}

.python, .r {
  background-color: #f8f9fa;
  border: none;
  border-radius: 0;
  padding: 20px;
}

table {
  margin: auto;
  border-top: 1px solid #666;
  border-bottom: 1px solid #666;
}

.figure {
  display: flex;
  justify-content: center;
}

table thead th { border-bottom: 1px solid #ddd; }
th, td { padding: 5px; }
thead, tfoot, tr:nth-child(even) { background: #eee; }
```

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(warning = TRUE, message = TRUE)
```

# Experiment
In this experiment we will rerun the FOP analysis using the provided dataset and scripts. The goal is to replicate the results and ensure the analysis is reproducible.

# Packages and Directory Setup
Below we will install the necessary R packages if they are not already installed.

```{r Installing Packages, echo=TRUE, results=FALSE, class.output="code", message=FALSE}
# Install the necessary packages if not already installed
base_packages <- c(
    # Data handling and Preprocessing
    "readr", "data.table", "dplyr", "tidyr", "tibble", "stringr", "Hmisc", "impute",

    # Visualization
    "ggplot2", "ggpubr", "ggprism", "plotly",
    "ComplexHeatmap", "ggVennDiagram", "ggtext",

    # Utilities
    "broom", "progress", "DT", "shiny", "pbkrtest", "emmeans", "afex", "rstudioapi", "furrr", "future",
    "progress",

    # Metagenomics
    "phyloseq", "vegan", "purrr", "taxize"
)

if (length(setdiff(base_packages, rownames(installed.packages()))) > 0) {
    install.packages(setdiff(base_packages, rownames(installed.packages())))
}

# Load Bioconductor packages
bioc_packages <- c("pheatmap", "DESeq2")
if (length(setdiff(bioc_packages, rownames(installed.packages()))) > 0) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
    }
    BiocManager::install(setdiff(bioc_packages, rownames(installed.packages())))
}

total_packages <- c(base_packages, bioc_packages)
invisible(lapply(total_packages, function(pkg) {
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

## Directory Setup
Next we will set up the directory structure for the analysis. Data was extracted from the Wynton server and placed in the `data` folder. Here, there are the following directories: 
* `Kneaddata` - contains the Kneaddata output files (useful for determining the number of host reads)
* `Kraken2` - contains the Kraken2 report files
* `Bracken` - contains the relative abundance files generated from Bracken
* `Humann` - contains the functional profiling files generated from Humann
In the following functional analysis, we will be using the Humann output files.

```{r Setting Directory Path, echo = TRUE}
# Set the working directory to the current directory
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Setting the path to the data and the file name
data_path <- "../data"
humann_path <- file.path(data_path, "Humann")
output_path <- "../results"

# Create the output directory if it doesn't exist
if (!dir.exists(output_path)) {
    dir.create(output_path, recursive = TRUE)
}
```

# Importing Data
First, we import the metadata to determine the samples and their respective treatments, timepoints, and groups.

```{r Importing Metadata, echo = TRUE}
# Import metadata
metadata <- read.csv(file.path(data_path, "Hsiao_shotgun_mapping_v3_final.csv"), header = TRUE, stringsAsFactors = FALSE) %>%
    dplyr::rename(SampleID = Sample.name) %>%
    column_to_rownames("SampleID")

# Import KEGG KO classification table
ko_details <- read_csv(file.path(data_path, "Overall-KEGG_KO_classification_table_Hsiao_FOP_metagenomics.csv"))
```

Next, we will import the Humann output files. Here we will import the compiled KO abundance table which is prenormalized to CPMs.

```{r Importing Humann Data, echo = TRUE}
# Import Humann KO abundance table
humann_ko_file <- file.path(humann_path, "all_samples_ko_cpm.tsv")
humann_ko_data <- read_tsv(humann_ko_file)
```

Next, we will clean the dataset and filter out low-abundance features and samples. Our thresholds will be to retain samples with at least 10% detected features and retain features present in at least 10% of samples.

```{r Cleaning Humann Data, echo = TRUE}
# Clean column names
df <- humann_ko_data %>%
  rename_with(~ gsub("_kneaddata_combined_Abundance-CPM", "", .x)) %>%
  rename("# Gene Family" = "KO")

# Remove all of the KOs with "|" and "UNMAPPED" or "UNGROUPED" 
df <- df %>%
  filter(!str_detect(KO, "\\||UNMAPPED|UNGROUPED")) %>%
  column_to_rownames("KO")

# Remove excess suffixes from sample names (column names)
colnames(df) <- stringr::word(colnames(df), 1, -3, sep = "_")

# Remap the column names based on the following mapping
remap <- c(
  "1_B11" = "FOP_02",
  "1_C11" = "CTRL_Abx_08",
  "1_D11" = "CTRL_Abx_09",
  "1_E11" = "CTRL_Abx_13",
  "1_F11" = "FOP_Abx_03",
  "1_G11" = "FOP_Abx_08"
)
colnames(df) <- ifelse(colnames(df) %in% names(remap), remap[colnames(df)], colnames(df))

# Reorganize the data alphabetically by sample names
df <- df %>% select(sort(colnames(.)))

# Run the filtering steps: retain samples with at least 10% detected features and retain features present in at least 10% of samples
df <- df %>%
  select(where(~ mean(. > 0) >= 0.1)) %>%
  filter(rowSums(. > 0) / ncol(.) >= 0.1)
```

## Creating Phyloseq Object
Next, we will create a phyloseq object to facilitate downstream analysis. This will involve creating an OTU table from the KO abundance data and integrating the metadata. Additionally, we will utilize the KO classification table to create a taxonomic table for the phyloseq object. This KO classification table has a depth of 4 level represented as the following in descending order: KO_Class, KO_Subclass, KO_Subclass_2, and KO_name. We will be utilizing KO_Subclass_2 for our differential abundance analysis. However, for the volcano plot visualization, we will be labeling via KO_subclass. 

```{r Creating Phyloseq Object, echo = TRUE}
# OTU table (KOs as rows)
otu <- phyloseq::otu_table(as.matrix(df), taxa_are_rows = TRUE)

# Taxonomy table
# Ensure KO column exists in ko_details
tax <- ko_details %>%
  dplyr::filter(KO %in% rownames(df)) %>%
  column_to_rownames("KO")

# Convert to matrix
tax <- phyloseq::tax_table(as.matrix(tax))

# Sample metadata
meta <- metadata %>%
  dplyr::filter(rownames(metadata) %in% colnames(df)) %>%
  phyloseq::sample_data()

# Construct phyloseq object
ps <- phyloseq(otu, tax, meta)
```

Here we will create an aggregated phyloseq object where the KOs are aggregated at the KO_Subclass_2 level by summing the counts of KOs that belong to the same KO_Subclass_2.

```{r Aggregation}
# Convert OTU table to data frame
otu_df <- as.data.frame(otu_table(ps))
if (!taxa_are_rows(ps)) {
  otu_df <- t(otu_df)
}

# Add taxonomy information
tax_df <- as.data.frame(tax_table(ps)) %>%
  rownames_to_column("KO")

# Merge counts with taxonomy
merged_df <- tax_df %>%
  select(KO, KO_Subclass_2) %>%
  left_join(otu_df %>% rownames_to_column("KO"), by = "KO")

# Aggregate counts by KO_Subclass_2
agg_df <- merged_df %>%
  group_by(KO_Subclass_2) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) %>%
  ungroup()

# Convert back to matrix (KO_Subclass_2 as rownames)
agg_mat <- as.matrix(agg_df[, -1])
rownames(agg_mat) <- agg_df$KO_Subclass_2

# Create new taxonomy table (still includes KO_Subclass_2 and higher levels)
agg_tax <- tax_df %>%
  distinct(KO_Subclass_2, .keep_all = TRUE) %>%
  filter(KO_Subclass_2 %in% rownames(agg_mat)) %>%
  column_to_rownames("KO_Subclass_2") %>%
  as.matrix() %>%
  tax_table()

# Build the new phyloseq object
ps_agg <- phyloseq(
  otu_table(agg_mat, taxa_are_rows = TRUE),
  agg_tax,
  sample_data(ps)
)
```

# Functional Analysis
With the phyloseq object created, we can now proceed with the functional analysis. This will include differential abundance testing and visualizations via volcano plots and heatmaps.

## Differential Abundance Testing
We will perform differential abundance testing using DESeq2 to identify KOs that are significantly different between the following groups:
  1. FOP (FOP_t1) vs WT (aka CTRL_t1)
  2. FOP (FOP_t1 + FOP_t2 + FOP_ABX_t1) vs WT (CTRL_t1 + CTRL_t2 + CTRL_abx_t1)
  3. FOP (FOP_ABX_t1) vs FOP GMA (FOP_ABX_t2)
  4. WT (CTRL_ABX_t1) vs WT GMA (Ctrl_abx_t2)

### FOP vs WT
Here, we are comparing FOP (FOP_t1 + FOP_t2 + FOP_ABX_t1) vs WT (CTRL_t1 + CTRL_t2 + CTRL_abx_t1). To do so, we will utilize the `Group2` variable in the metadata which precomputes these groupings. We will first run the differential abundance testing at the KO level (non-aggregated).

```{r Differential Abundance Testing, echo = TRUE}
# Filter phyloseq object to only include relevant samples
ps_filtered <- prune_samples(sample_data(ps)$Type %in% c("CTRL_ABX_T1", "CTRL_T1", "CTRL_T2","FOP_ABX_T1",  "FOP_T1",  "FOP_T2"), ps)
ps_filtered <- filter_taxa(
  ps_filtered,
  function(x) sum(x > 0) > (0.3 * length(x)),
  prune = TRUE
)

# Convert to DESeq2 object
sample_data(ps_filtered)$Group2 <- factor(
  as.character(sample_data(ps_filtered)$Group2),
  levels = c("CTRL", "FOP")
)
dds <- phyloseq_to_deseq2(ps_filtered, ~ Group2)

# Run DESeq2
dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds, fitType = "local", sfType = "poscounts")
phy_deseq_1_res <- results(dds, name = "Group2_FOP_vs_CTRL", pAdjustMethod = "fdr")

# Merge results with KO details
phy_deseq_1_res_df <- as.data.frame(phy_deseq_1_res) %>%
  rownames_to_column("KO") %>%
  left_join(ko_details, by = "KO") %>%
  filter(!is.na(padj))

# Print the number of significant KOs based on an adjusted p-value threshold of 0.05
significant_kos <- phy_deseq_1_res_df %>% filter(padj < 0.05)
if (nrow(significant_kos) == 0) {
  cat("No significant KOs found (adjusted p-value < 0.05).\n")
} else {
  cat("Number of significant KOs (adjusted p-value < 0.05):", nrow(significant_kos), "\n")
}
```

```{r Volcano Plot FOP vs WT, echo = TRUE, fig.width=12, fig.height=6}
# Prepare volcano data
df <- phy_deseq_1_res_df

# Stabilize padj and compute -log10
eps <- .Machine$double.xmin
df$padj_fix <- ifelse(
  is.na(df$padj) | df$padj <= 0,
  eps,
  df$padj
)
df$neglog10_padj <- -log10(df$padj_fix)

# Axis caps to avoid compression from outliers
y_cap <- max(
  stats::quantile(df$neglog10_padj, 0.99, na.rm = TRUE),
  3
)

x_cap <- max(
  stats::quantile(abs(df$log2FoldChange), 0.99, na.rm = TRUE),
  1.5
)

# Volcano plot
vp1 <- ggplot(
  df,
  aes(x = log2FoldChange, y = neglog10_padj)
) +
  geom_point(
    color = "grey70",
    alpha = 0.75,
    size = 2.5
  ) +
  geom_point(
    data = subset(df, padj_fix < 0.05),
    aes(color = KO_Subclass),
    size = 2.8
  ) +
  geom_vline(
    xintercept = 0,
    color = "grey40",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    color = "red",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  labs(
    title = "Differential Abundance Volcano Plot",
    x = expression("Log"[2] * " Fold Change (FOP vs WT)"),
    y = expression("-Log"[10] * "(adj p-value)"),
    color = "KO Subclass"
  ) +
  coord_cartesian(
    xlim = c(-x_cap, x_cap),
    ylim = c(0, y_cap)
  ) +
  theme_prism(base_size = 16, border = TRUE) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 14),
    legend.text = element_text(face = "italic", size = 12),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )

print(vp1)

# Save to Figures folder
ggsave(
  filename = file.path(fig_dir, "Volcano_FOP_vs_WT.pdf"),
  plot = vp1,
  device = cairo_pdf,
  width = 12,
  height = 6,
  dpi = 300
)

ggsave(
  filename = file.path(fig_dir, "Volcano_FOP_vs_WT.tiff"),
  plot = vp1,
  device = "tiff",
  compression = "lzw",
  width = 12,
  height = 6,
  dpi = 300
)
```


```{r Pooled FOP_vs_WT Heatmap Averaged, echo=TRUE, fig.width=8, fig.height=8}
# Significant KOs from pooled FOP vs WT model
sig_kos <- phy_deseq_1_res_df %>%
  dplyr::filter(!is.na(padj), padj < 0.05) %>%
  dplyr::arrange(padj)

cat("Significant KOs (padj < 0.05):", nrow(sig_kos), "\n")

if (nrow(sig_kos) == 0) {
  stop("No significant KOs; cannot draw heatmap.")
}

if (!"KO" %in% names(sig_kos)) {
  stop("phy_deseq_1_res_df must contain a column named 'KO'.")
}

# Normalized counts from DESeq object
nc <- DESeq2::counts(dds, normalized = TRUE)

sig_ids <- intersect(sig_kos$KO, rownames(nc))
nc_sig <- nc[sig_ids, , drop = FALSE]

sig_kos <- sig_kos %>%
  dplyr::filter(KO %in% sig_ids)

nc_sig <- nc_sig[sig_kos$KO, , drop = FALSE]

# Sample metadata and type renaming
md <- as.data.frame(sample_data(ps_filtered))

md$TypeRenamed <- md$Type
md$TypeRenamed <- gsub("^CTRL", "WT", md$TypeRenamed)
md$TypeRenamed <- gsub("ABX", "GMA", md$TypeRenamed)

keep_types <- c(
  "WT_T1", "WT_T2", "WT_GMA_T1",
  "FOP_T1", "FOP_T2", "FOP_GMA_T1"
)

keep_samples <- rownames(md)[md$TypeRenamed %in% keep_types]
keep_samples <- intersect(keep_samples, colnames(nc_sig))

nc_sig <- nc_sig[, keep_samples, drop = FALSE]
md <- md[keep_samples, , drop = FALSE]

# Average across TypeRenamed
mat_long <- as.data.frame(t(nc_sig))
mat_long$TypeRenamed <- md$TypeRenamed

mat_avg <- mat_long %>%
  dplyr::group_by(TypeRenamed) %>%
  dplyr::summarise(dplyr::across(where(is.numeric), mean)) %>%
  as.data.frame()

rownames(mat_avg) <- mat_avg$TypeRenamed
mat_avg <- mat_avg[, -1, drop = FALSE]

# KOs as rows, Types as columns
mat_avg <- t(mat_avg)

desired_order <- c(
  "WT_T1", "WT_T2", "WT_GMA_T1",
  "FOP_T1", "FOP_T2", "FOP_GMA_T1"
)
desired_order <- desired_order[desired_order %in% colnames(mat_avg)]
mat_avg <- mat_avg[, desired_order, drop = FALSE]

# Log transform and row z-score
mat_log <- log10(mat_avg + 1)
mat_z <- t(scale(t(mat_log)))

# Remove rows that cannot be scaled
keep_rows <- apply(mat_z, 1, function(x) all(is.finite(x)))
mat_z <- mat_z[keep_rows, , drop = FALSE]
sig_kos <- sig_kos[keep_rows, , drop = FALSE]

if (nrow(mat_z) == 0) {
  stop("No rows remain after filtering non-finite values.")
}

# Row labels: KO_Name if available, else KO ID
row_labels <- ifelse(
  "KO_Name" %in% names(sig_kos) & !is.na(sig_kos$KO_Name) & sig_kos$KO_Name != "",
  sig_kos$KO_Name,
  sig_kos$KO
)
rownames(mat_z) <- row_labels

# Row annotation: KO_Subclass
if ("KO_Subclass" %in% names(sig_kos)) {
  ann_row <- data.frame(KO_Subclass = sig_kos$KO_Subclass)
  rownames(ann_row) <- rownames(mat_z)

  subclass_lvls <- sort(unique(stats::na.omit(ann_row$KO_Subclass)))
  subclass_cols <- setNames(
    scales::hue_pal()(length(subclass_lvls)),
    subclass_lvls
  )

  ann_colors <- list(
    KO_Subclass = subclass_cols,
    Group = c(WT = "#999999", FOP = "#000000")
  )
} else {
  ann_row <- NULL
  ann_colors <- list(
    Group = c(WT = "#999999", FOP = "#000000")
  )
}

# Column annotation
ann_col <- data.frame(
  Type = colnames(mat_z),
  Group = ifelse(grepl("^WT", colnames(mat_z)), "WT", "FOP")
)
rownames(ann_col) <- colnames(mat_z)

# Plot heatmap
pal <- colorRampPalette(c("#2166AC", "white", "#B2182B"))

pheatmap::pheatmap(
  mat_z,
  color = pal(101),
  breaks = seq(-3, 3, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_row = ann_row,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 8,
  fontsize_col = 15,
  angle_col = 45,
  border_color = NA,
  main = "Significant KOs (FOP vs WT pool, averaged by Type)"
)

# Save heatmap to Figures folder
ht <- grid::grid.grab()

grDevices::pdf(
  file = file.path(fig_dir, "Pooled_FOP_vs_WT_KO_heatmap.pdf"),
  width = 10,
  height = 12,
  useDingbats = FALSE
)
grid::grid.draw(ht)
dev.off()

grDevices::tiff(
  file = file.path(fig_dir, "Pooled_FOP_vs_WT_KO_heatmap.tiff"),
  width = 10,
  height = 12,
  units = "in",
  res = 300,
  compression = "lzw"
)
grid::grid.draw(ht)
dev.off()

# Optional: save matrix used in heatmap
write.csv(
  mat_z,
  file = file.path(fig_dir, "Pooled_FOP_vs_WT_KO_heatmap_matrix.csv")
)
```