---
title: "FOP Analysis Rerun"
author: "Authors: Ethan Dinh"
date: "2025-10-2"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: '5'
    df_print: paged
  html_notebook:
    theme: spacelab
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes 
      smooth_scroll: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
---

```{css, echo=FALSE}
img {
  align-self: center;
  height: fit-content;
}

pre {
 margin: 20px 0;
}

.python, .r {
  background-color: #f8f9fa;
  border: none;
  border-radius: 0;
  padding: 20px;
}

table {
  margin: auto;
  border-top: 1px solid #666;
  border-bottom: 1px solid #666;
}

.figure {
  display: flex;
  justify-content: center;
}

table thead th { border-bottom: 1px solid #ddd; }
th, td { padding: 5px; }
thead, tfoot, tr:nth-child(even) { background: #eee; }
```

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(warning = TRUE, message = TRUE)
```

# Experiment
In this experiment we will rerun the FOP analysis using the provided dataset and scripts. The goal is to replicate the results and ensure the analysis is reproducible.

# Packages and Directory Setup
Below we will install the necessary R packages if they are not already installed.

```{r Installing Packages, echo=TRUE, results=FALSE, class.output="code", message=FALSE}
# Install the necessary packages if not already installed
base_packages <- c(
    # Data handling and Preprocessing
    "readr", "data.table", "dplyr", "tidyr", "tibble", "stringr", "Hmisc", "impute",

    # Visualization
    "ggplot2", "ggpubr", "ggprism", "plotly",
    "ComplexHeatmap", "ggVennDiagram", "ggtext",

    # Utilities
    "broom", "progress", "DT", "shiny", "pbkrtest", "emmeans", "afex", "rstudioapi", "furrr", "future",
    "progress",

    # Metagenomics
    "phyloseq", "vegan", "purrr", "taxize"
)

if (length(setdiff(base_packages, rownames(installed.packages()))) > 0) {
    install.packages(setdiff(base_packages, rownames(installed.packages())))
}

# Load Bioconductor packages
bioc_packages <- c("pheatmap", "DESeq2")
if (length(setdiff(bioc_packages, rownames(installed.packages()))) > 0) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
    }
    BiocManager::install(setdiff(bioc_packages, rownames(installed.packages())))
}

total_packages <- c(base_packages, bioc_packages)
invisible(lapply(total_packages, function(pkg) {
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

## Directory Setup
Next we will set up the directory structure for the analysis. Data was extracted from the Wynton server and placed in the `data` folder. Here, there are the following directories: 
* `Kneaddata` - contains the Kneaddata output files (useful for determining the number of host reads)
* `Kraken2` - contains the Kraken2 report files
* `Bracken` - contains the relative abundance files generated from Bracken

```{r Setting Directory Path, echo = TRUE}
# Set the working directory to the current directory
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Setting the path to the data and the file name
data_path <- "../data"
kraken_path <- file.path(data_path, "Kraken2")
bracken_path <- file.path(data_path, "BrackenComplete")
kneaddata_path <- file.path(data_path, "Kneaddata")
output_path <- "../results"

# Create the output directory if it doesn't exist
if (!dir.exists(output_path)) {
    dir.create(output_path, recursive = TRUE)
}
```

# Importing Data
To generate Bray-Curtis dissimilarity plots, we need to import the relative abundance files from Bracken. Since each sample has a separate file, we will read in all of the files and combine them into a single data frame.

First, we import the metadata to determine the samples and their respective treatments, timepoints, and groups.

```{r Importing Metadata, echo = TRUE}
# Import metadata
metadata <- read.csv(file.path(data_path, "Hsiao_shotgun_mapping_v3_final.csv"), header = TRUE, stringsAsFactors = FALSE) %>%
    dplyr::rename(SampleID = Sample.name) %>%
    column_to_rownames("SampleID")
```

Next, we import the Bracken data. We will read in all of the files in the Bracken directory that end with `_S.tsv`, extract the sample name from the file name, and combine them into a single data frame. We will also relabel the samples to match the metadata.

```{r Importing Bracken Data, echo = TRUE}
# Define input directory and taxonomy levels
bracken_dir <- file.path(data_path, "BrackenComplete")
levels <- c("D", "K", "P", "C", "O", "F", "G", "S")

# Read and combine all Bracken files into one table per level
read_bracken_level <- function(level) {
    files <- list.files(bracken_dir, pattern = paste0("_bracken_", level, "\\.tsv$"), full.names = TRUE)
    df <- map_dfr(files, function(f) {
        sample_id <- str_replace(basename(f), paste0("_bracken_", level, "\\.tsv$"), "")
        read_tsv(f, show_col_types = FALSE) %>%
          select(name, taxonomy_id, new_est_reads, fraction_total_reads) %>%
          mutate(Sample = sample_id, Level = level)
        })
    return(df)
}

bracken_all <- map_dfr(levels, read_bracken_level) %>%
    dplyr::mutate(Sample = stringr::word(Sample, 1, -3, sep = "_"))

# Relabeling Samples
bracken_all$Sample <- gsub("1_B11", "FOP_02", bracken_all$Sample)
bracken_all$Sample <- gsub("1_C11", "CTRL_Abx_08", bracken_all$Sample)
bracken_all$Sample <- gsub("1_D11", "CTRL_Abx_09", bracken_all$Sample)
bracken_all$Sample <- gsub("1_E11", "CTRL_Abx_13", bracken_all$Sample)
bracken_all$Sample <- gsub("1_F11", "FOP_Abx_03", bracken_all$Sample)
bracken_all$Sample <- gsub("1_G11", "FOP_Abx_08", bracken_all$Sample)

# Separate treatment from replicate number. The replicate is the last item after the underscore and the treatment is everything before it. Keep both.
bracken_all <- bracken_all %>%
    dplyr::mutate(sampleID = Sample) %>%
    separate(Sample, into = c("Treatment", "Replicate"), sep = "_(?=[^_]+$)")

# Reshape Bracken Data to Wide Format
bracken_wide <- bracken_all %>%
    select(taxonomy_id, sampleID, new_est_reads) %>%
    pivot_wider(
        names_from = sampleID,
        values_from = new_est_reads,
        values_fill = 0
    )

# Select the species level data
bracken_species <- bracken_wide %>%
    filter(taxonomy_id %in% bracken_all$taxonomy_id[bracken_all$Level == "S"])

# Create OTU table
otu_table_df <- bracken_species %>%
    column_to_rownames("taxonomy_id") %>%
    as.matrix()
```

## Creating Taxonomic Tree
To create a taxonomic tree, we need to retrieve the taxonomic lineage for each species in our OTU table. We will use the `taxize` package to query the NCBI taxonomy database. This allows us to trace each species back to its higher taxonomic ranks (e.g., genus, family, order, etc.). When running this code, note that NCBI may throttle requests if too many are made in a short period of time. To avoid this, we will implement caching to store previously retrieved taxonomic lineages. Additionally, a delay will be added between requests to further reduce the risk of throttling.

```{r Creating Taxonomic Tree, echo = TRUE, warning=FALSE}
# --- Extract taxids ---
species_taxids <- rownames(otu_table_df)
species_taxids <- as.numeric(species_taxids)

# --- Config ---
cache_file <- file.path(data_path, "taxonomy_cache.rds")

if (file.exists(cache_file)) {
    message("Loading cached taxonomy results...")
    lineage_list <- readRDS(cache_file)
    done <- names(lineage_list)
} else {
    message("Querying NCBI taxonomy via taxize...")
    lineage_list <- list()
    done <- character(0)
}

# Ensure that all of the done taxids have data and not NULL
for (tid in done) {
    if (is.null(lineage_list[[tid]][1])) {
        done <- setdiff(done, tid)
    }
}

# Checking for the remaining taxids after cache load
remaining <- setdiff(as.character(species_taxids), done)
if (length(remaining) > 0) {
    for (tid in remaining) {
        Sys.sleep(0.2)
        res <- tryCatch(
            classification(tid, db = "ncbi"),
            error = function(e) {
              message("Skipping taxid ", tid, ": ", e$message)
              NULL
            }
        )
        lineage_list[[as.character(tid)]] <- res
        saveRDS(lineage_list, cache_file)
    }
    
    message("Finished taxonomy retrieval. Saved to: ", cache_file)
} else {
    message("All taxids were found in cache.")
}
```
Once we have the taxonomy data retreived, we need to reshape it into Hierarchical Table format.

```{r Reshaping Kraken Data, echo = TRUE}
taxonomy_table <- map_dfr(names(lineage_list), function(tid) {
    x <- lineage_list[[tid]][[1]]  # unwrap the inner data.frame
    
    if (is.null(x) || !"rank" %in% names(x)) return(NULL)
    
    # Extract each level if it exists
    tibble(
        taxonomy_id = tid,
        Kingdom = x$name[x$rank %in% c("superkingdom", "domain")][1],
        Phylum  = x$name[x$rank == "phylum"][1],
        Class   = x$name[x$rank == "class"][1],
        Order   = x$name[x$rank == "order"][1],
        Family  = x$name[x$rank == "family"][1],
        Genus   = x$name[x$rank == "genus"][1],
        Species = x$name[x$rank == "species"][1]
    )
})

taxonomy_mat <- taxonomy_table %>%
    column_to_rownames("taxonomy_id") %>%
    as.matrix()
```

Now that we have the taxonomy information, we can merge it with the Bracken data to create a complete data frame that includes species names, estimated reads, treatment groups, and replicate numbers.

```{r Merging Bracken and Taxonomy Data, echo = TRUE}
common_ids <- intersect(rownames(otu_table_df), rownames(taxonomy_mat))
otu_table_df <- otu_table_df[common_ids, , drop = FALSE]
taxonomy_mat <- taxonomy_mat[common_ids, , drop = FALSE]
```

## Creating Phyloseq Object
To perform downstream analyses, we will create a Phyloseq object using the OTU table we just created. We will also create a sample metadata table to include treatment information.

```{r Creating Phyloseq Object, echo = TRUE}
ps <- phyloseq(
    otu_table(otu_table_df, taxa_are_rows = TRUE),
    tax_table(taxonomy_mat),
    sample_data(metadata)
)

# Filter low-abundance taxa
ps <- filter_taxa(ps, function(x) sum(x > 0) > (0.1 * length(x)), TRUE)
```

# Data Normalization
Before performing any analyses, we need to normalize the data to account for differences in sequencing depth across samples. We will use VST (Variance Stabilizing Transformation) from the DESeq2 package for normalization.

```{r Data Normalization, echo = TRUE, message=FALSE}
ps_vst <- ps %>% tax_glom(taxrank = "Species")
ps_deseq2 <- phyloseq_to_deseq2(ps_vst, ~1)
ps_deseq2_norm <- estimateSizeFactors(ps_deseq2, type = "poscounts")
otu_table(ps_vst) <- otu_table(counts(ps_deseq2_norm, normalized = TRUE), taxa_are_rows = TRUE)
```

#QC Plots

```{r QC plots, echo = TRUE, message=FALSE}
# Helper: 
clean_names_simple <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9]+", "_", x)
  x <- gsub("^_|_$", "", x)
  x
}

# Paths
knead_dir <- kneaddata_path
map_path  <- file.path(data_path, "Hsiao_shotgun_mapping_v3_final.csv")

# Read metadata
meta <- read_csv(map_path, show_col_types = FALSE)
names(meta) <- clean_names_simple(names(meta))

stopifnot("sample_name" %in% names(meta))

meta <- meta %>%
  mutate(sample_name = as.character(sample_name))

# Read one kneaddata read_count file
read_knead_readcount <- function(fp) {
  x <- read_tsv(fp, col_names = FALSE, show_col_types = FALSE)

  # Row 1 contains headers
  hdr <- as.character(unlist(x[1, ]))
  hdr <- clean_names_simple(hdr)

  # Row 2 contains values
  vals <- x[2, ]
  out <- as_tibble(vals)
  names(out) <- hdr

  if (!"sample" %in% names(out)) {
    stop("No 'sample' column detected after parsing: ", fp)
  }

  out <- out %>%
    mutate(sample = as.character(sample)) %>%
    mutate(across(-sample, ~ suppressWarnings(as.numeric(.x))))

  out
}

# Load all kneaddata files
files <- list.files(
  knead_dir,
  pattern = "kneaddata\\.read_count\\.tsv$",
  full.names = TRUE
)

knead_wide <- purrr::map_dfr(files, read_knead_readcount)

# Kneaddata sample names look like CTRL_01_S3_L003, metadata uses CTRL_01
knead_wide <- knead_wide %>%
  mutate(sample_name = gsub("_S\\d+_L\\d+$", "", sample))

# Join to metadata
qc <- knead_wide %>%
  left_join(meta, by = "sample_name")

cat("\n--- JOIN CHECK ---\n")
cat("kneaddata samples:", n_distinct(knead_wide$sample_name), "\n")
cat("metadata samples:", n_distinct(meta$sample_name), "\n")
cat("joined samples w/ metadata:", sum(!is.na(qc$group)), "\n")

# Create QC summary table
qc_slim <- qc %>%
  transmute(
    sample_name,
    group = group,
    input_reads   = raw_pair1,
    trimmed_reads = trimmed_pair1,
    final_reads   = final_pair1,
    pct_removed   = 100 * (input_reads - final_reads) / input_reads
  ) %>%
  filter(!is.na(input_reads) & !is.na(final_reads))
```
```{r shotgun-qc, echo=TRUE, message=TRUE, warning=TRUE, fig.width=8, fig.height=6, fig.show="hold"}
# Figures output folder
fig_dir <- "Figures"

if (!dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}

# Plot: Kneaddata read counts by stage
qc_long <- qc_slim %>%
  pivot_longer(
    cols = c(input_reads, trimmed_reads, final_reads),
    names_to = "stage",
    values_to = "reads"
  ) %>%
  mutate(
    stage = factor(
      stage,
      levels = c("input_reads", "trimmed_reads", "final_reads"),
      labels = c("Input", "Trimmed", "Final")
    )
  )

p_reads <- ggplot(qc_long, aes(x = sample_name, y = reads, fill = stage)) +
  geom_col() +
  labs(
    title = "Kneaddata Read Counts by Stage",
    x = "Sample",
    y = "Reads",
    fill = "Stage"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
    axis.ticks.x = element_blank()
  )

print(p_reads)

# Save figure
ggsave(
  filename = file.path(fig_dir, "QC_Kneaddata_Reads_By_Stage.pdf"),
  plot = p_reads,
  width = 10,
  height = 5,
  device = "pdf"
)
```

# Data Analysis

## Taxonomic diversity analysis 
Here we will calculate and plot the alpha and beta diversity metrics.

### Alpha Diversity
Here we calculate and plot the alpha diversity metrics (Richness and Shannon Diversity Index) for each sample group. We will use boxplots to visualize the differences between groups and perform statistical tests to assess significance.

```{r Alpha Diversity, echo = TRUE, message=FALSE, fig.width=12, fig.height=6}
# Build diversity dataframe
df <- data.frame(sample_data(ps_vst))
df$Richness <- vegan::specnumber(t(otu_table(ps_vst)))
df$Shannon  <- vegan::diversity(t(otu_table(ps_vst)), index = "shannon")

# Pretty labels
label_map <- c(
  "CTRL_T1"     = "WT t1",
  "CTRL_T2"     = "WT t2",
  "FOP_T1"      = "FOP t1",
  "FOP_T2"      = "FOP t2",
  "CTRL_ABX_T1" = "WT GMA t1",
  "CTRL_ABX_T2" = "WT GMA t2",
  "FOP_ABX_T1"  = "FOP GMA t1",
  "FOP_ABX_T2"  = "FOP GMA t2"
)

df$TypePretty <- factor(
  label_map[as.character(df$Type)],
  levels = c(
    "WT t1", "WT t2",
    "FOP t1", "FOP t2",
    "WT GMA t1", "WT GMA t2",
    "FOP GMA t1", "FOP GMA t2"
  )
)

# Helpers
  safe_wilcox <- function(y, g) {
  g <- droplevels(factor(g))
  if (length(unique(na.omit(g))) != 2) return(NA_real_)
  tryCatch(wilcox.test(y ~ g, exact = FALSE)$p.value, error = function(e) NA_real_)
}

  n12 <- function(g) {
  tt <- table(droplevels(factor(g)))
  vals <- as.integer(tt)
  if (length(vals) == 1) vals <- c(vals, NA_integer_)
  names(vals) <- c("n1", "n2")
  vals
}

# Selected Wilcoxon comparisons
# WT (t1+t2) vs FOP (t1+t2), non-ABX only
nonabx_idx <- df$Type %in% c("CTRL_T1", "CTRL_T2", "FOP_T1", "FOP_T2")
pool_g <- ifelse(df$Type[nonabx_idx] %in% c("CTRL_T1", "CTRL_T2"), "WT", "FOP")

p_pool_R <- safe_wilcox(df$Richness[nonabx_idx], pool_g)
p_pool_S <- safe_wilcox(df$Shannon[nonabx_idx],  pool_g)
n_pool   <- n12(pool_g)

# WT GMA t1 vs WT GMA t2
wtg_idx <- df$TypePretty %in% c("WT GMA t1", "WT GMA t2")
p_wtg_R <- safe_wilcox(df$Richness[wtg_idx], df$TypePretty[wtg_idx])
p_wtg_S <- safe_wilcox(df$Shannon[wtg_idx],  df$TypePretty[wtg_idx])
n_wtg   <- n12(df$TypePretty[wtg_idx])

# FOP GMA t1 vs FOP GMA t2
fopg_idx <- df$TypePretty %in% c("FOP GMA t1", "FOP GMA t2")
p_fopg_R <- safe_wilcox(df$Richness[fopg_idx], df$TypePretty[fopg_idx])
p_fopg_S <- safe_wilcox(df$Shannon[fopg_idx],  df$TypePretty[fopg_idx])
n_fopg   <- n12(df$TypePretty[fopg_idx])

# Assemble stats table
pvals_tbl <- tibble::tibble(
  Metric     = c("Richness", "Shannon", "Richness", "Shannon", "Richness", "Shannon"),
  Comparison = c(
    "WT (t1+t2) vs FOP (t1+t2)",
    "WT (t1+t2) vs FOP (t1+t2)",
    "WT GMA t1 vs WT GMA t2",
    "WT GMA t1 vs WT GMA t2",
    "FOP GMA t1 vs FOP GMA t2",
    "FOP GMA t1 vs FOP GMA t2"
  ),
  Group1     = c("WT", "WT", "WT GMA t1", "WT GMA t1", "FOP GMA t1", "FOP GMA t1"),
  Group2     = c("FOP", "FOP", "WT GMA t2", "WT GMA t2", "FOP GMA t2", "FOP GMA t2"),
  n1         = c(n_pool["n1"], n_pool["n1"], n_wtg["n1"], n_wtg["n1"], n_fopg["n1"], n_fopg["n1"]),
  n2         = c(n_pool["n2"], n_pool["n2"], n_wtg["n2"], n_wtg["n2"], n_fopg["n2"], n_fopg["n2"]),
  p_value    = c(p_pool_R, p_pool_S, p_wtg_R, p_wtg_S, p_fopg_R, p_fopg_S)
) %>%
  dplyr::mutate(
    p_value = as.numeric(p_value),
    p_sig = dplyr::case_when(
      is.na(p_value)  ~ "NA",
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ "ns"
    ),
    p_pretty = ifelse(is.na(p_value), "NA", paste0("p = ", signif(p_value, 3)))
  )

print(pvals_tbl %>% dplyr::select(Metric, Comparison, Group1, Group2, n1, n2, p_pretty, p_sig))

# Sample sizes by displayed group
n_table <- df %>%
  dplyr::group_by(TypePretty) %>%
  dplyr::summarise(
    n_samples = dplyr::n(),
    n_mice = dplyr::n_distinct(MouseID),
    .groups = "drop"
  )

print(n_table)

# Descriptive statistics by displayed group
desc_stats <- df %>%
  dplyr::group_by(TypePretty) %>%
  dplyr::summarise(
    n_samples     = dplyr::n(),
    n_mice        = dplyr::n_distinct(MouseID),
    richness_mean = mean(Richness, na.rm = TRUE),
    richness_sd   = sd(Richness, na.rm = TRUE),
    shannon_mean  = mean(Shannon, na.rm = TRUE),
    shannon_sd    = sd(Shannon, na.rm = TRUE),
    .groups = "drop"
  )

print(desc_stats)

# Colors
fill_vals <- c("#F8766D", "#00BFC4", "#7CAE00", "#C77CFF")
df$Group <- factor(df$Group, levels = c("CTRL", "FOP", "CTRL_ABX", "FOP_ABX"))

# LME calculations
AD_lme_R <- anova(lmer(Richness ~ as.factor(Type) + (1 | MouseID), data = df))
AD_lme_S <- anova(lmer(Shannon  ~ as.factor(Type) + (1 | MouseID), data = df))

print(AD_lme_R)
print(AD_lme_S)


# Plot spacing
maxR <- max(df$Richness, na.rm = TRUE)
maxS <- max(df$Shannon,  na.rm = TRUE)
y_pad_R <- maxR * 0.08
y_pad_S <- maxS * 0.08

# Richness plot
plot_AD1 <- ggplot(df, aes(x = TypePretty, y = Richness, fill = Group)) +
  geom_boxplot(width = 0.6, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, height = 0),
    size = 1.6,
    alpha = 0.6
  ) +
  scale_fill_manual(values = fill_vals, name = "Group") +
  labs(y = "Taxon Richness", x = "") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.3))) +
  annotate(
    "text",
    x = 7.5,
    y = maxR + 4 * y_pad_R,
    label = paste("ANOVA (LME)\nP =", signif(AD_lme_R$`Pr(>F)`[1], 3))
  ) +
  theme_prism(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.margin = margin(10, 30, 10, 10)
  ) +
  coord_cartesian(clip = "off")

# Shannon plot
plot_AD2 <- ggplot(df, aes(x = TypePretty, y = Shannon, fill = Group)) +
  geom_boxplot(width = 0.6, outlier.shape = NA) +
  geom_point(
    position = position_jitter(width = 0.15, height = 0),
    size = 1.6,
    alpha = 0.6
  ) +
  scale_fill_manual(values = fill_vals, name = "Group") +
  labs(y = "Shannon Diversity Index", x = "") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.3))) +
  annotate(
    "text",
    x = 2.0,
    y = maxS + 4 * y_pad_S,
    label = paste("ANOVA (LME)\nP =", signif(AD_lme_S$`Pr(>F)`[1], 3))
  ) +
  theme_prism(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.margin = margin(10, 30, 10, 10)
  ) +
  coord_cartesian(clip = "off")

# Combine plots
plt <- ggpubr::ggarrange(
  plot_AD1,
  plot_AD2,
  common.legend = TRUE,
  legend = "bottom"
)

print(plt)

# Save figure
ggsave(
  filename = file.path(fig_dir, "Alpha_Diversity_Boxplots.pdf"),
  plot = plt,
  width = 10,
  height = 6,
  device = cairo_pdf
)

ggsave(
  filename = file.path(fig_dir, "Alpha_Diversity_Boxplots.tiff"),
  plot = plt,
  width = 10,
  height = 6,
  units = "in",
  dpi = 300,
  compression = "lzw"
)

# Save stats outputs
write.csv(
  pvals_tbl,
  file = file.path(fig_dir, "Alpha_Diversity_Selected_Wilcoxon_Tests.csv"),
  row.names = FALSE
)

write.csv(
  n_table,
  file = file.path(fig_dir, "Alpha_Diversity_Sample_Sizes.csv"),
  row.names = FALSE
)

write.csv(
  desc_stats,
  file = file.path(fig_dir, "Alpha_Diversity_Descriptive_Stats.csv"),
  row.names = FALSE
)

write.csv(
  as.data.frame(AD_lme_R),
  file = file.path(fig_dir, "Alpha_Diversity_LME_Richness_ANOVA.csv"),
  row.names = TRUE
)

write.csv(
  as.data.frame(AD_lme_S),
  file = file.path(fig_dir, "Alpha_Diversity_LME_Shannon_ANOVA.csv"),
  row.names = TRUE
)
```

### Beta Diversity via Bray-Curtis PCoA
To determine the differences in microbial community composition between groups, we will calculate the Bray-Curtis dissimilarity and perform Principal Coordinates Analysis (PCoA). We will visualize the results using scatter plots and assess statistical significance using PERMANOVA.

#### WT compared to FOP Mice after GMA 
Here we are comparing the Bray-Curtis PCoA of WT and FOP mice after GMA treatment.

Once we have calculated the Bray-Curtis PCoA, we need to isolate the first two axes and add them to the sample metadata for plotting. 

```{r Plotting, echo = TRUE, message=FALSE, fig.width=8, fig.height=6}
# Prepare metadata
sdf <- data.frame(sample_data(ps_vst), check.names = FALSE)

# Infer timepoint if missing
if (!"Timepoint" %in% names(sdf)) {
  tp_from_name <- sub(".*_T([12]).*", "T\\1", rownames(sdf))
  tp_from_name[!grepl("_T[12]", rownames(sdf))] <- NA
  sdf$Timepoint <- tp_from_name
}
sdf$Timepoint <- factor(sdf$Timepoint, levels = c("T1", "T2"))

# Pretty labels
sdf$Type <- as.character(sdf$Type)
sdf$TypePretty <- dplyr::recode(
  sdf$Type,
  "CTRL_T1"     = "WT t1",
  "CTRL_T2"     = "WT t2",
  "FOP_T1"      = "FOP t1",
  "FOP_T2"      = "FOP t2",
  "CTRL_ABX_T1" = "WT GMA t1",
  "CTRL_ABX_T2" = "WT GMA t2",
  "FOP_ABX_T1"  = "FOP GMA t1",
  "FOP_ABX_T2"  = "FOP GMA t2"
)

sdf$TypePretty <- factor(
  sdf$TypePretty,
  levels = c(
    "WT t1", "WT t2",
    "FOP t1", "FOP t2",
    "WT GMA t1", "WT GMA t2",
    "FOP GMA t1", "FOP GMA t2"
  )
)

# 4-color plotting groups
sdf$GroupPlot <- dplyr::case_when(
  sdf$Type %in% c("CTRL_T1", "CTRL_T2") ~ "WT",
  sdf$Type %in% c("FOP_T1", "FOP_T2") ~ "FOP",
  sdf$Type %in% c("CTRL_ABX_T1", "CTRL_ABX_T2") ~ "WT GMA",
  sdf$Type %in% c("FOP_ABX_T1", "FOP_ABX_T2") ~ "FOP GMA",
  TRUE ~ NA_character_
)

sdf$GroupPlot <- factor(
  sdf$GroupPlot,
  levels = c("WT GMA", "WT", "FOP", "FOP GMA")
)

# Before / After GMA grouping
sdf$GMATiming <- dplyr::case_when(
  sdf$Type %in% c("CTRL_ABX_T2", "FOP_ABX_T2") ~ "After GMA",
  sdf$Type %in% c(
    "CTRL_T1", "CTRL_T2",
    "FOP_T1", "FOP_T2",
    "CTRL_ABX_T1", "FOP_ABX_T1"
  ) ~ "Before GMA",
  TRUE ~ NA_character_
)

sdf$GMATiming <- factor(
  sdf$GMATiming,
  levels = c("Before GMA", "After GMA")
)

# WT ellipse
sdf$WTEllipse <- ifelse(
  sdf$Type %in% c("CTRL_T1", "CTRL_T2", "CTRL_ABX_T1"),
  "WT ellipse",
  NA
)

# FOP ellipse
sdf$FOPEllipse <- ifelse(
  sdf$Type %in% c("FOP_T1", "FOP_T2", "FOP_ABX_T1"),
  "FOP ellipse",
  NA
)

sample_data(ps_vst) <- sample_data(sdf)

# Subset to desired groups
phy_plot <- subset_samples(
  ps_vst,
  Type %in% c(
    "CTRL_T1", "CTRL_T2",
    "FOP_T1", "FOP_T2",
    "CTRL_ABX_T1", "CTRL_ABX_T2",
    "FOP_ABX_T1", "FOP_ABX_T2"
  )
)

# Bray-Curtis PCoA
pcoa_bray <- ordinate(
  phy_plot,
  method = "PCoA",
  distance = "bray"
)

scores_df <- as.data.frame(
  pcoa_bray$vectors[, 1:2, drop = FALSE]
)

colnames(scores_df) <- c("Axis.1", "Axis.2")
scores_df$SampleID <- rownames(scores_df)

plot_meta <- data.frame(
  sample_data(phy_plot),
  check.names = FALSE
)

plot_meta$SampleID <- rownames(plot_meta)

plot_df <- dplyr::left_join(
  scores_df,
  plot_meta,
  by = "SampleID"
)

pc1_var <- if (!is.null(pcoa_bray$values$Relative_eig)) {
  round(100 * pcoa_bray$values$Relative_eig[1], 1)
} else { NA }

pc2_var <- if (!is.null(pcoa_bray$values$Relative_eig)) {
  round(100 * pcoa_bray$values$Relative_eig[2], 1)
} else { NA }

# Report
cat("\n================ BETA DIVERSITY REPORT ================\n")
cat("Distance metric: Bray-Curtis\n")
cat("Ordination: PCoA\n")
cat("All samples retained (no manual outlier removal).\n")
cat("=======================================================\n\n")

plot_n_table <- data.frame(sample_data(phy_plot), check.names = FALSE) %>%
  dplyr::group_by(TypePretty, GroupPlot, GMATiming) %>%
  dplyr::summarise(
    n_samples = dplyr::n(),
    n_mice = dplyr::n_distinct(MouseID),
    .groups = "drop"
  )

print(plot_n_table)

# Aesthetics
color_map <- c(
  "WT GMA"  = "#20C4C9",
  "WT"      = "#F8766D",
  "FOP"     = "#7CAE00",
  "FOP GMA" = "#B276E8"
)

shape_map <- c(
  "Before GMA" = 16,
  "After GMA"  = 17
)

xr <- range(plot_df$Axis.1, na.rm = TRUE)
yr <- range(plot_df$Axis.2, na.rm = TRUE)

# Plot
plot_bray <- ggplot(plot_df, aes(x = Axis.1, y = Axis.2)) +

  geom_point(
    aes(color = GroupPlot, shape = GMATiming),
    size = 6
  ) +

  stat_ellipse(
    data = subset(plot_df, !is.na(WTEllipse)),
    aes(group = WTEllipse),
    type = "t",
    linewidth = 2.2,
    color = "#F8766D",
    linetype = "dashed"
  ) +

  stat_ellipse(
    data = subset(plot_df, !is.na(FOPEllipse)),
    aes(group = FOPEllipse),
    type = "t",
    linewidth = 2.2,
    color = "#7CAE00",
    linetype = "dashed"
  ) +

  stat_ellipse(
    data = subset(plot_df, GMATiming == "Before GMA"),
    aes(group = GMATiming),
    type = "t",
    linewidth = 2.2,
    color = "black",
    linetype = "solid"
  ) +

  stat_ellipse(
    data = subset(plot_df, GMATiming == "After GMA"),
    aes(group = GMATiming),
    type = "t",
    linewidth = 2.2,
    color = "black",
    linetype = "dashed"
  ) +

  scale_color_manual(values = color_map, name = "Group") +
  scale_shape_manual(values = shape_map, guide = "none") +

  labs(
    x = paste0("Principal Coordinate 1 (", pc1_var, "%)"),
    y = paste0("Principal Coordinate 2 (", pc2_var, "%)")
  ) +

  annotate(
    "text",
    x = xr[2] - 0.01 * diff(xr),
    y = yr[2] - 0.01 * diff(yr),
    label = "Before GMA",
    hjust = 1,
    vjust = 1,
    size = 10
  ) +

  annotate(
    "text",
    x = xr[1] + 0.03 * diff(xr),
    y = yr[1] + 0.35 * diff(yr),
    label = "After GMA",
    hjust = 0,
    vjust = 0,
    size = 10
  ) +

  theme_bw(base_size = 20) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 3),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_text(size = 28),
    axis.title.y = element_text(size = 28),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 17),
    legend.position = c(0.12, 0.18),
    legend.justification = c(0, 0),
    legend.background = element_blank(),
    legend.key = element_blank()
  )

print(plot_bray)

# Save outputs
ggsave(
  file.path(fig_dir, "BrayCurtis_PCoA_WT_FOP_GMA.pdf"),
  plot = plot_bray,
  width = 8,
  height = 6,
  device = cairo_pdf
)

ggsave(
  file.path(fig_dir, "BrayCurtis_PCoA_WT_FOP_GMA.tiff"),
  plot = plot_bray,
  width = 8,
  height = 6,
  units = "in",
  dpi = 300,
  compression = "lzw"
)

write.csv(
  plot_n_table,
  file.path(fig_dir, "BrayCurtis_PCoA_Group_Sample_Sizes.csv"),
  row.names = FALSE
)
```

```{r PERMANOVA P Values, echo=TRUE, message=FALSE, fig.width=8, fig.height=6}
# Helper 
perm_test <- function(phy, dist_method = "bray", rhs, title) {
  set.seed(123)

  d  <- phyloseq::distance(phy, method = dist_method)
  md <- as(sample_data(phy), "data.frame")

  form <- stats::as.formula(paste("d ~", rhs))
  res  <- vegan::adonis2(form, data = md)

  cat(sprintf(
    "%s (%s): F=%0.5f, R^2=%0.5f, p=%s\n",
    title,
    dist_method,
    res$F[1],
    res$R2[1],
    signif(res$`Pr(>F)`[1], 3)
  ))

  invisible(res)
}
# bray-curtis PERMANOVA
# CTRL_T1 vs FOP_T1
phy_ct1_ft1 <- prune_samples(
  sample_data(ps_vst)$Type %in% c("CTRL_T1","FOP_T1"),
  ps_vst
)
perm_test(phy_ct1_ft1, "bray", "Type", "CTRL_T1 vs FOP_T1")

# CTRL_T1 vs CTRL_ABX_T1
phy_ct1_cabt1 <- prune_samples(
  sample_data(ps_vst)$Type %in% c("CTRL_T1","CTRL_ABX_T1"),
  ps_vst
)
perm_test(phy_ct1_cabt1, "bray", "Type", "CTRL_T1 vs CTRL_ABX_T1")

# FOP_T1 vs FOP_ABX_T1
phy_ft1_fabt1 <- prune_samples(
  sample_data(ps_vst)$Type %in% c("FOP_T1","FOP_ABX_T1"),
  ps_vst
)
perm_test(phy_ft1_fabt1, "bray", "Type", "FOP_T1 vs FOP_ABX_T1")

# CTRL_T2 vs CTRL_ABX_T2
phy_ct2_cabt2 <- prune_samples(
  sample_data(ps_vst)$Type %in% c("CTRL_T2","CTRL_ABX_T2"),
  ps_vst
)
perm_test(phy_ct2_cabt2, "bray", "Type", "CTRL_T2 vs CTRL_ABX_T2")

# FOP_T2 vs FOP_ABX_T2
phy_ft2_fabt2 <- prune_samples(
  sample_data(ps_vst)$Type %in% c("FOP_T2","FOP_ABX_T2"),
  ps_vst
)
perm_test(phy_ft2_fabt2, "bray", "Type", "FOP_T2 vs FOP_ABX_T2")

# CTRL_T1 + CTRL_ABX_T1 vs FOP_T1 + FOP_ABX_T1
phy_cabt1_fabt1 <- prune_samples(
  sample_data(ps_vst)$Type %in% c(
    "CTRL_T1","CTRL_ABX_T1",
    "FOP_T1","FOP_ABX_T1"
  ),
  ps_vst
)
perm_test(
  phy_cabt1_fabt1,
  "bray",
  "Group2",
  "CTRL_T1+CTRL_ABX_T1 vs FOP_T1+FOP_ABX_T1"
)

# CTRL_T1 + CTRL_T2 + CTRL_ABX_T1 vs FOP_T1 + FOP_T2 + FOP_ABX_T1
phy_cabt1_fabt1 <- prune_samples(
  sample_data(ps_vst)$Type %in% c(
    "CTRL_T1","CTRL_T2","CTRL_ABX_T1",
    "FOP_T1","FOP_T2","FOP_ABX_T1"
  ),
  ps_vst
)
perm_test(
  phy_cabt1_fabt1,
  "bray",
  "Group2",
  "CTRL_T1+CTRL_T2+CTRL_ABX_T1 vs FOP_T1+FOP_T2+FOP_ABX_T1"
)
```

# Relative Abundance Plots
Here we will create stacked bar plots to visualize the relative abundance of taxa at different taxonomic levels (Kingdom, Phylum, Class) across the experimental groups.

```{r Taxonomic Bar Plots, echo=TRUE, message=FALSE}
# Helper function for stacked barplots
plot_taxa_bar <- function(ps_obj, taxrank, outfile, facet_levels, y_break = NULL, save = TRUE) {
    
    # Validate object
    if (!inherits(ps_obj, "phyloseq")) stop("ps_obj must be a phyloseq object.")
    if (!taxrank %in% rank_names(ps_obj)) {
        stop(paste("Tax rank", taxrank, "not found in object. Available ranks:",
                   paste(rank_names(ps_obj), collapse = ", ")))
    }
    
    # Remove taxa with zero counts
    ps_obj <- prune_taxa(taxa_sums(ps_obj) > 0, ps_obj)
    
    # Agglomerate to desired rank
    agg <- tax_glom(ps_obj, taxrank = taxrank)
    
    # Transform to relative abundance
    rel <- transform_sample_counts(agg, function(x) x / sum(x))
    
    # Build plot
    p <- plot_bar(rel, fill = taxrank) +
    geom_bar(aes_string(color = taxrank, fill = taxrank),
             stat = "identity", position = "stack") +
    xlab("") +
    ylab("Relative Abundance (%)") +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    facet_grid(~factor(Type, levels = facet_levels), scales = "free_x", space = "free_x") +
    theme_bw(base_size = 14) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
        axis.title.y = element_text(size = 14, face = "bold"),
        legend.text = element_text(face = "italic", size = 10),
        legend.title = element_text(face = "bold", size = 11),
        strip.text.x = element_text(size = 11, face = "bold")
    )
    
    # Save figure
    if (save)
        ggsave(filename = outfile, plot = p, width = 12, height = 5, dpi = 300)
    
    return(p)
}

# Facet order (experimental groups)
facet_levels <- c(
    "CTRL_T1", "CTRL_T2",
    "FOP_T1", "FOP_T2",
    "CTRL_ABX_T1", "CTRL_ABX_T2",
    "FOP_ABX_T1", "FOP_ABX_T2"
)
```

## Generate and Save Taxonomic Bar Plots
Here we will generate stacked bar plots showing the relative abundance of taxa at the Kingdom and Phylum levels across different experimental groups.

```{r Generating Taxonomic Bar Plots, echo=TRUE, message=FALSE, fig.width=12, fig.height=5}
# Generate plots

library(scales)

ps_vst <- ps
ps_deseq2 <- phyloseq_to_deseq2(ps_vst, ~1)
ps_deseq2_norm <- estimateSizeFactors(ps_deseq2, type = "poscounts")
otu_table(ps_vst) <- otu_table(counts(ps_deseq2_norm, normalized = TRUE), taxa_are_rows = TRUE)

domain_plot <- plot_taxa_bar(
    ps_obj = ps_vst,
    taxrank = "Kingdom",
    outfile = "Figures/FOP-Kraken-taxonomy_domain_level_barplot.pdf",
    facet_levels = facet_levels,
    save = FALSE
)

phylum_plot <- plot_taxa_bar(
    ps_obj = ps_vst,
    taxrank = "Phylum",
    outfile = "Figures/FOP-Kraken-taxonomy_phylum_level_barplot.pdf",
    facet_levels = facet_levels,
    save = FALSE
)

class_plot <- plot_taxa_bar(
    ps_obj = ps_vst,
    taxrank = "Class",
    outfile = "Figures/FOP-Kraken-taxonomy_class_level_barplot.pdf",
    facet_levels = facet_levels,
    save = FALSE
)

# Display
domain_plot
phylum_plot
class_plot
```

```{r Taxonomic Bar Plots for paper, echo=TRUE, message=FALSE}
# Helper function for stacked barplots
plot_taxa_bar <- function(ps_obj, taxrank, facet_levels) {

  if (!inherits(ps_obj, "phyloseq")) {
    stop("ps_obj must be a phyloseq object.")
  }

  if (!taxrank %in% rank_names(ps_obj)) {
    stop(
      paste(
        "Tax rank", taxrank,
        "not found in object. Available ranks:",
        paste(rank_names(ps_obj), collapse = ", ")
      )
    )
  }

  # Remove taxa with zero counts
  ps_obj <- prune_taxa(
    taxa_sums(ps_obj) > 0,
    ps_obj
  )

  # Agglomerate to desired rank
  agg <- tax_glom(
    ps_obj,
    taxrank = taxrank
  )

  # Relative abundance
  rel <- transform_sample_counts(
    agg,
    function(x) x / sum(x)
  )

  # Plot
  p <- plot_bar(rel, fill = taxrank) +
    geom_bar(
      aes_string(color = taxrank, fill = taxrank),
      stat = "identity",
      position = "stack"
    ) +
    xlab("") +
    ylab("Relative Abundance (%)") +
    scale_y_continuous(
      labels = percent_format(accuracy = 1)
    ) +
    facet_grid(
      ~ factor(Type, levels = facet_levels),
      scales = "free_x",
      space = "free_x"
    ) +
    theme_bw(base_size = 14) +
    theme(
      axis.text.x = element_text(
        angle = 60,
        hjust = 1,
        vjust = 1,
        size = 11
      ),
      axis.title.y = element_text(
        size = 14,
        face = "bold"
      ),
      legend.text = element_text(
        face = "italic",
        size = 10
      ),
      legend.title = element_text(
        face = "bold",
        size = 11
      ),
      strip.text.x = element_text(
        size = 12,
        face = "bold"
      )
    )

  return(p)
}

# Facet order
facet_levels <- c(
  "WT_T1", "WT_T2",
  "FOP_T1", "FOP_T2",
  "WT_GMA_T1", "WT_GMA_T2",
  "FOP_GMA_T1", "FOP_GMA_T2"
)

# Normalize + relabel
ps_vst <- ps

ps_deseq2 <- phyloseq_to_deseq2(
  ps_vst,
  ~1
)

ps_deseq2_norm <- estimateSizeFactors(
  ps_deseq2,
  type = "poscounts"
)

otu_table(ps_vst) <- otu_table(
  counts(ps_deseq2_norm, normalized = TRUE),
  taxa_are_rows = TRUE
)

# Fix sample names
sn <- sample_names(ps_vst)

sn <- gsub("^CTRL_", "WT_", sn)
sn <- gsub("Abx", "GMA", sn)

sample_names(ps_vst) <- sn

# Fix metadata labels
sd <- data.frame(sample_data(ps_vst))

sd$Type <- as.character(sd$Type)
sd$Type <- gsub("CTRL_ABX", "WT_GMA", sd$Type)
sd$Type <- gsub("FOP_ABX", "FOP_GMA", sd$Type)
sd$Type <- gsub("CTRL", "WT", sd$Type)

sample_data(ps_vst) <- sample_data(sd)

# Generate plots
domain_plot <- plot_taxa_bar(
  ps_obj = ps_vst,
  taxrank = "Kingdom",
  facet_levels = facet_levels
)

phylum_plot <- plot_taxa_bar(
  ps_obj = ps_vst,
  taxrank = "Phylum",
  facet_levels = facet_levels
)

class_plot <- plot_taxa_bar(
  ps_obj = ps_vst,
  taxrank = "Class",
  facet_levels = facet_levels
)

# Show in Rmd
domain_plot
phylum_plot
class_plot

# Save to Figures folder

# DOMAIN
ggsave(
  filename = file.path(
    fig_dir,
    "FOP-Kraken-taxonomy_domain_level_barplot.pdf"
  ),
  plot = domain_plot,
  width = 16,
  height = 7,
  units = "in",
  dpi = 300
)

ggsave(
  filename = file.path(
    fig_dir,
    "FOP-Kraken-taxonomy_domain_level_barplot.tiff"
  ),
  plot = domain_plot,
  width = 16,
  height = 7,
  units = "in",
  dpi = 300,
  compression = "lzw"
)

# PHYLUM
ggsave(
  filename = file.path(
    fig_dir,
    "FOP-Kraken-taxonomy_phylum_level_barplot.pdf"
  ),
  plot = phylum_plot,
  width = 20,
  height = 8,
  units = "in",
  dpi = 300
)

ggsave(
  filename = file.path(
    fig_dir,
    "FOP-Kraken-taxonomy_phylum_level_barplot.tiff"
  ),
  plot = phylum_plot,
  width = 20,
  height = 8,
  units = "in",
  dpi = 300,
  compression = "lzw"
)

# CLASS
ggsave(
  filename = file.path(
    fig_dir,
    "FOP-Kraken-taxonomy_class_level_barplot.pdf"
  ),
  plot = class_plot,
  width = 20,
  height = 8,
  units = "in",
  dpi = 300
)

ggsave(
  filename = file.path(
    fig_dir,
    "FOP-Kraken-taxonomy_class_level_barplot.tiff"
  ),
  plot = class_plot,
  width = 20,
  height = 8,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```
```{r Revert WT_GMA_to_CTRL_ABX, echo=TRUE, message=FALSE}
# This assumes ps_vst currently has WT / WT_GMA / FOP_GMA names
# and you want to go back to CTRL / CTRL_ABX / FOP_ABX

# 1) Revert sample names
sn <- sample_names(ps_vst)

# Do the more specific patterns first, then the general ones
sn <- gsub("WT_GMA_",  "CTRL_ABX_", sn)
sn <- gsub("FOP_GMA_", "FOP_ABX_",  sn)
sn <- gsub("^WT_",     "CTRL_",     sn)

sample_names(ps_vst) <- sn

# (Optional) also revert ps if you think you changed it earlier
# sn_ps <- sample_names(ps)
# sn_ps <- gsub("WT_GMA_",  "CTRL_ABX_", sn_ps)
# sn_ps <- gsub("FOP_GMA_", "FOP_ABX_",  sn_ps)
# sn_ps <- gsub("^WT_",     "CTRL_",     sn_ps)
# sample_names(ps) <- sn_ps

# 2) Revert Type metadata (facet / group labels)
sd <- data.frame(sample_data(ps_vst))

sd$Type <- as.character(sd$Type)

# Specific patterns first
sd$Type <- gsub("WT_GMA_T1",  "CTRL_ABX_T1", sd$Type)
sd$Type <- gsub("WT_GMA_T2",  "CTRL_ABX_T2", sd$Type)
sd$Type <- gsub("FOP_GMA_T1", "FOP_ABX_T1",  sd$Type)
sd$Type <- gsub("FOP_GMA_T2", "FOP_ABX_T2",  sd$Type)

# Then the baseline ones
sd$Type <- gsub("WT_T1", "CTRL_T1", sd$Type)
sd$Type <- gsub("WT_T2", "CTRL_T2", sd$Type)

sample_data(ps_vst) <- sample_data(sd)

# Quick sanity check
table(sample_data(ps_vst)$Type)
```


# Differential Abundance Analysis
Here we will perform differential abundance analysis at the phylum level using DESeq2, comparing the combined FOP groups (FOP_ABX_T1, FOP_T1, FOP_T2) against the combined CTRL groups (CTRL_ABX_T1, CTRL_T1, CTRL_T2). We will also create a volcano plot to visualize the results.

## Phylum-Level Differential Abundance and Volcano Plot
We first want to prepare the data for DESeq2 analysis by constructing a phylum-level count matrix from phyloseq object. Then, we will perform the differential abundance analysis and create a volcano plot to visualize significant changes. In specific, we will be comparing the FOP groups (FOP_ABX_T1, FOP_T1, FOP_T2) against the CTRL groups (CTRL_ABX_T1, CTRL_T1, CTRL_T2), both without antibiotic treatment.

```{r Dataset Preparation all FOP vs CTRL for DESeq2, echo=TRUE, message=FALSE}
ps_filtered <- prune_samples(sample_data(ps_vst)$Type %in% c("CTRL_ABX_T1", "CTRL_T1", "CTRL_T2","FOP_ABX_T1",  "FOP_T1",  "FOP_T2"), ps_vst)

# Filter taxa present in over 30% of the samples
ps_filtered <- filter_taxa(ps_filtered, function(x) sum(x > 0) > (0.3*length(x)), TRUE)

# Convert to factor
sample_data(ps_filtered)$Type <- factor(sample_data(ps_filtered)$Type)

sample_data(ps_filtered)$Group <- factor(sample_data(ps_filtered)$Group)

# Convert to DESeq2 object
phy_deseq <- phyloseq_to_deseq2(ps_filtered, ~ Group)
```

```{r Differential Expression Analysis all FOP vs CTRL, echo=TRUE, message=FALSE, fig.width=8, fig.height=6}
# --- Run DESeq2 analysis ---
phy_deseq_1 <- DESeq(phy_deseq, fitType = "local", sfType = "poscounts")

# Extract results for the specified comparison
phy_deseq_1_res <- results(phy_deseq_1,
                           name = "Group_FOP_vs_CTRL",
                           pAdjustMethod = "fdr")

# Convert to data frame and preserve OTU/ASV names
phy_deseq_1_res_df <- as.data.frame(phy_deseq_1_res) %>%
  tibble::rownames_to_column("FeatureID")

# Extract taxonomy table and prepare for merge
tax_df <- as.data.frame(tax_table(ps_filtered)) %>%
  tibble::rownames_to_column("FeatureID")

# Merge DESeq2 results with taxonomy information
phy_deseq_1_res_cft1 <- dplyr::inner_join(phy_deseq_1_res_df, tax_df, by = "FeatureID")

# Replace NA adjusted p-values with 1 (non-significant)
phy_deseq_1_res_cft1$padj[is.na(phy_deseq_1_res_cft1$padj)] <- 1

# Subset to significant features (padj < 0.05)
phy_deseq_1_res_cft1_sig <- dplyr::filter(phy_deseq_1_res_cft1, padj < 0.05)

# Sanity check
print("Significant features: ", nrow(phy_deseq_1_res_cft1_sig))
```

```{r Volcano Plot all FOP vs CTRL, echo=TRUE, message=FALSE, fig.width=8, fig.height=6}
# Create Figures folder
fig_dir <- "Figures"
if (!dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}

# Volcano plot
df <- phy_deseq_1_res_cft1

# Stabilize padj and compute -log10
eps <- .Machine$double.xmin
df$padj_fix <- ifelse(is.na(df$padj) | df$padj <= 0, eps, df$padj)
df$neglog10_padj <- -log10(df$padj_fix)

# Axis caps to reduce compression from extreme outliers
y_cap <- max(stats::quantile(df$neglog10_padj, 0.99, na.rm = TRUE), 8)
x_cap <- max(stats::quantile(abs(df$log2FoldChange), 0.99, na.rm = TRUE), 3)

vp1 <- ggplot(df, aes(x = log2FoldChange, y = neglog10_padj)) +
  geom_point(color = "grey70", alpha = 0.75, size = 2.5) +
  geom_point(
    data = subset(df, padj_fix < 0.05),
    aes(color = Phylum),
    size = 2.8
  ) +
  geom_vline(xintercept = 0, color = "grey40", linetype = "dashed", linewidth = 0.7) +
  geom_hline(yintercept = -log10(0.05), color = "red", linetype = "dashed", linewidth = 0.7) +
  labs(
    title = "Differential Abundance Volcano Plot",
    x = expression("Log"[2] * " Fold Change (FOP vs CTRL)"),
    y = expression("-Log"[10] * "(p-value)"),
    color = "Phylum"
  ) +
  coord_cartesian(xlim = c(-x_cap, x_cap), ylim = c(0, y_cap)) +
  theme_prism(base_size = 16, border = TRUE) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position  = "right",
    legend.title     = element_text(size = 14),
    legend.text      = element_text(face = "italic", size = 12),
    plot.title       = element_text(face = "bold", hjust = 0.5),
    axis.title.x     = element_text(face = "bold"),
    axis.title.y     = element_text(face = "bold")
  )

print(vp1)

# Save figure
ggsave(
  filename = file.path(fig_dir, "Volcano_FOP_vs_CTRL_T1_T2_ABXT1.pdf"),
  plot = vp1,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  device = cairo_pdf
)

ggsave(
  filename = file.path(fig_dir, "Volcano_FOP_vs_CTRL_T1_T2_ABXT1.tiff"),
  plot = vp1,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

```{r Dataset Preparation FOP_ABX_T2_vs_FOP_Pool, echo=TRUE, message=FALSE}
# Keep only FOP-related groups (T1, T2, ABX_T1, ABX_T2)
ps_fop_abx2 <- prune_samples(
  sample_data(ps_vst)$Type %in% c("FOP_T1", "FOP_T2", "FOP_ABX_T1", "FOP_ABX_T2"),
  ps_vst
)

# Filter taxa present in >30% of samples
ps_fop_abx2 <- filter_taxa(
  ps_fop_abx2,
  function(x) sum(x > 0) > (0.3 * length(x)),
  prune = TRUE
)

# Define comparison factor:
#   - "FOP_POOL" = FOP_T1, FOP_T2, FOP_ABX_T1
#   - "FOP_ABX_T2" = FOP_ABX_T2 (comparison group)
grp <- ifelse(
  sample_data(ps_fop_abx2)$Type == "FOP_ABX_T2",
  "FOP_ABX_T2",
  "FOP_POOL"
)

sample_data(ps_fop_abx2)$GroupFOPAbx2 <- factor(
  grp,
  levels = c("FOP_POOL", "FOP_ABX_T2")  # FOP_POOL as reference
)

# Build DESeq2 object
dds_fop_abx2 <- phyloseq_to_deseq2(ps_fop_abx2, ~ GroupFOPAbx2)
```

```{r Differential Expression FOP_ABX_T2_vs_FOP_Pool, echo=TRUE, message=FALSE}
dds_fop_abx2 <- DESeq(dds_fop_abx2, fitType = "local", sfType = "poscounts")

# Contrast: FOP_ABX_T2 vs FOP_POOL (log2FC > 0 => higher in FOP_ABX_T2)
res_fop_abx2 <- results(
  dds_fop_abx2,
  contrast = c("GroupFOPAbx2", "FOP_ABX_T2", "FOP_POOL"),
  pAdjustMethod = "fdr"
)

# Preserve feature IDs
res_fop_abx2_df <- as.data.frame(res_fop_abx2) %>%
  tibble::rownames_to_column("FeatureID")

# Bring in taxonomy
tax_fop_abx2 <- as.data.frame(tax_table(ps_fop_abx2)) %>%
  tibble::rownames_to_column("FeatureID")

# Merge results + taxonomy
res_fop_abx2_merged <- dplyr::inner_join(res_fop_abx2_df, tax_fop_abx2, by = "FeatureID")

# Replace NA padj with 1 (non-significant)
res_fop_abx2_merged$padj[is.na(res_fop_abx2_merged$padj)] <- 1

message(sprintf("Significant features (padj < 0.05): %d",
                sum(res_fop_abx2_merged$padj < 0.05, na.rm = TRUE)))
```

```{r Volcano Plot FOP_ABX_T2_vs_FOP_Pool, echo=TRUE, message=FALSE, fig.width=8, fig.height=6}
# Volcano plot
vp_fop_abx2 <- ggplot(
  res_fop_abx2_merged,
  aes(x = log2FoldChange, y = -log10(padj))
) +

  # All points
  geom_point(
    color = "grey70",
    alpha = 0.7,
    size = 3
  ) +

  # Significant points
  {
    if ("Phylum" %in% names(res_fop_abx2_merged)) {
      geom_point(
        data = subset(res_fop_abx2_merged, padj < 0.05),
        aes(color = Phylum),
        size = 3
      )
    } else {
      geom_point(
        data = subset(res_fop_abx2_merged, padj < 0.05),
        color = "black",
        size = 3
      )
    }
  } +

  # Threshold lines
  geom_vline(
    xintercept = 0,
    color = "grey40",
    linetype = "dashed",
    linewidth = 0.8
  ) +

  geom_hline(
    yintercept = -log10(0.05),
    color = "red",
    linetype = "dashed",
    linewidth = 0.8
  ) +

  # Labels
  labs(
    title = "Differential Abundance: FOP_ABX_T2 vs FOP Pool (T1 + T2 + ABX_T1)",
    x = expression("Log"[2] * " Fold Change (FOP_ABX_T2 vs FOP pool)"),
    y = expression("-Log"[10] * "(adj. p-value)"),
    color = "Phylum"
  ) +

  theme_prism(base_size = 16, border = TRUE) +

  theme(
    legend.position = "right",
    legend.title = element_text(size = 14),
    legend.text = element_text(face = "italic", size = 12),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )

print(vp_fop_abx2)

# Save to Figures folder
fig_dir <- "Figures"

if (!dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}

ggsave(
  filename = file.path(
    fig_dir,
    "Volcano_FOP_ABX_T2_vs_FOP_Pool.pdf"
  ),
  plot = vp_fop_abx2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  device = cairo_pdf
)

ggsave(
  filename = file.path(
    fig_dir,
    "Volcano_FOP_ABX_T2_vs_FOP_Pool.tiff"
  ),
  plot = vp_fop_abx2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

```{r Dataset Preparation CTRL_ABX_T2_vs_CTRL_Pool, echo=TRUE, message=FALSE}
# Keep only the CTRL-related groups
ps_ctrl_abx2 <- prune_samples(
  sample_data(ps_vst)$Type %in% c("CTRL_T1", "CTRL_T2", "CTRL_ABX_T1", "CTRL_ABX_T2"),
  ps_vst
)

# Filter taxa present in >30% of samples
ps_ctrl_abx2 <- filter_taxa(
  ps_ctrl_abx2,
  function(x) sum(x > 0) > (0.3 * length(x)),
  prune = TRUE
)

# Define comparison factor: CTRL_POOL = T1 + T2 + ABX_T1, comparison = ABX_T2
grp <- ifelse(
  sample_data(ps_ctrl_abx2)$Type == "CTRL_ABX_T2",
  "CTRL_ABX_T2",
  "CTRL_POOL"
)

sample_data(ps_ctrl_abx2)$GroupCTRLAbx2 <- factor(
  grp,
  levels = c("CTRL_POOL", "CTRL_ABX_T2") # CTRL_POOL is reference
)

# Build DESeq2 object
dds_ctrl_abx2 <- phyloseq_to_deseq2(ps_ctrl_abx2, ~ GroupCTRLAbx2)
```

```{r Differential Expression CTRL_ABX_T2_vs_CTRL_Pool, echo=TRUE, message=FALSE}
dds_ctrl_abx2 <- DESeq(dds_ctrl_abx2, fitType = "local", sfType = "poscounts")

# Contrast: CTRL_ABX_T2 vs CTRL_POOL
res_ctrl_abx2 <- results(
  dds_ctrl_abx2,
  contrast = c("GroupCTRLAbx2", "CTRL_ABX_T2", "CTRL_POOL"),
  pAdjustMethod = "fdr"
)

# Preserve feature IDs
res_ctrl_abx2_df <- as.data.frame(res_ctrl_abx2) %>%
  tibble::rownames_to_column("FeatureID")

# Bring in taxonomy
tax_ctrl_abx2 <- as.data.frame(tax_table(ps_ctrl_abx2)) %>%
  tibble::rownames_to_column("FeatureID")

# Merge results + taxonomy
res_ctrl_abx2_merged <- dplyr::inner_join(res_ctrl_abx2_df, tax_ctrl_abx2, by = "FeatureID")

# Replace NA padj with 1 (non-significant)
res_ctrl_abx2_merged$padj[is.na(res_ctrl_abx2_merged$padj)] <- 1

message(sprintf(
  "Significant features (padj < 0.05): %d",
  sum(res_ctrl_abx2_merged$padj < 0.05, na.rm = TRUE)
))
```

```{r Volcano Plot CTRL_ABX_T2_vs_CTRL_Pool, echo=TRUE, message=FALSE, fig.width=8, fig.height=6}
vp_ctrl_abx2 <- ggplot(
  res_ctrl_abx2_merged,
  aes(x = log2FoldChange, y = -log10(padj))
) +

  # All points
  geom_point(
    color = "grey70",
    alpha = 0.7,
    size = 3
  ) +

  # Significant points colored by Phylum
  {
    if ("Phylum" %in% names(res_ctrl_abx2_merged)) {
      geom_point(
        data = subset(res_ctrl_abx2_merged, padj < 0.05),
        aes(color = Phylum),
        size = 3
      )
    } else {
      geom_point(
        data = subset(res_ctrl_abx2_merged, padj < 0.05),
        color = "black",
        size = 3
      )
    }
  } +

  # Threshold lines
  geom_vline(
    xintercept = 0,
    color = "grey40",
    linetype = "dashed",
    linewidth = 0.8
  ) +

  geom_hline(
    yintercept = -log10(0.05),
    color = "red",
    linetype = "dashed",
    linewidth = 0.8
  ) +

  # Labels
  labs(
    title = "Differential Abundance: CTRL_ABX_T2 vs CTRL Pool (T1 + T2 + ABX_T1)",
    x = expression("Log"[2] * " Fold Change (CTRL_ABX_T2 vs CTRL pool)"),
    y = expression("-Log"[10] * "(adj. p-value)"),
    color = "Phylum"
  ) +

  theme_prism(base_size = 16, border = TRUE) +

  theme(
    legend.position = "right",
    legend.title = element_text(size = 14),
    legend.text = element_text(face = "italic", size = 12),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )

print(vp_ctrl_abx2)

# Save to Figures folder
fig_dir <- "Figures"

if (!dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}

ggsave(
  filename = file.path(
    fig_dir,
    "Volcano_CTRL_ABX_T2_vs_CTRL_Pool.pdf"
  ),
  plot = vp_ctrl_abx2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  device = cairo_pdf
)

ggsave(
  filename = file.path(
    fig_dir,
    "Volcano_CTRL_ABX_T2_vs_CTRL_Pool.tiff"
  ),
  plot = vp_ctrl_abx2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```


```{r Harmonize_Phylum_Colors_Across_Volcanoes, echo=TRUE, message=FALSE}
# Create Figures folder
fig_dir <- "Figures"

if (!dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}

# Collect shared Phylum levels
all_phyla <- sort(unique(c(
  as.character(phy_deseq_1_res_cft1$Phylum),
  as.character(res_fop_abx2_merged$Phylum),
  as.character(res_ctrl_abx2_merged$Phylum)
)))

all_phyla <- all_phyla[!is.na(all_phyla) & all_phyla != ""]

# Shared color palette
phylum_cols <- setNames(
  scales::hue_pal()(length(all_phyla)),
  all_phyla
)

# Harmonize factor levels
phy_deseq_1_res_cft1$Phylum <- factor(
  phy_deseq_1_res_cft1$Phylum,
  levels = all_phyla
)

res_fop_abx2_merged$Phylum <- factor(
  res_fop_abx2_merged$Phylum,
  levels = all_phyla
)

res_ctrl_abx2_merged$Phylum <- factor(
  res_ctrl_abx2_merged$Phylum,
  levels = all_phyla
)

# Apply shared color scale
vp1 <- vp1 + scale_color_manual(
  values = phylum_cols,
  drop = FALSE
)

vp_fop_abx2 <- vp_fop_abx2 + scale_color_manual(
  values = phylum_cols,
  drop = FALSE
)

vp_ctrl_abx2 <- vp_ctrl_abx2 + scale_color_manual(
  values = phylum_cols,
  drop = FALSE
)

# Re-print
print(vp1)
print(vp_fop_abx2)
print(vp_ctrl_abx2)

# Save harmonized versions
ggsave(
  filename = file.path(fig_dir, "Volcano_FOP_vs_CTRL_T1_T2_ABXT1_same.pdf"),
  plot = vp1,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  device = cairo_pdf
)

ggsave(
  filename = file.path(fig_dir, "Volcano_FOP_vs_CTRL_T1_T2_ABXT1_same.tiff"),
  plot = vp1,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)

ggsave(
  filename = file.path(fig_dir, "Volcano_FOP_ABX_T2_vs_FOP_Pool_same.pdf"),
  plot = vp_fop_abx2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  device = cairo_pdf
)

ggsave(
  filename = file.path(fig_dir, "Volcano_FOP_ABX_T2_vs_FOP_Pool_same.tiff"),
  plot = vp_fop_abx2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)

ggsave(
  filename = file.path(fig_dir, "Volcano_CTRL_ABX_T2_vs_CTRL_Pool_same.pdf"),
  plot = vp_ctrl_abx2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  device = cairo_pdf
)

ggsave(
  filename = file.path(fig_dir, "Volcano_CTRL_ABX_T2_vs_CTRL_Pool_same.tiff"),
  plot = vp_ctrl_abx2,
  width = 7,
  height = 5,
  units = "in",
  dpi = 300,
  compression = "lzw"
)
```

```{r Heatmap_sig_species_WT_GMA_vs_WT_pool_AVERAGED, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Significant Feature IDs
sig_ids <- res_ctrl_abx2_merged %>%
  dplyr::filter(!is.na(padj), padj < 0.05) %>%
  dplyr::pull(FeatureID) %>%
  unique()

if (length(sig_ids) == 0) {
  stop("No significant features for WT GMA vs WT pool.")
}

# Normalized counts from DESeq object
nc <- counts(dds_ctrl_abx2, normalized = TRUE)
sig_ids <- intersect(sig_ids, rownames(nc))
nc_sig <- nc[sig_ids, , drop = FALSE]

# Map FeatureID to Species
tax_ctrl <- as.data.frame(tax_table(ps_ctrl_abx2))
tax_ctrl$FeatureID <- rownames(tax_ctrl)

species_vec <- tax_ctrl$Species[match(rownames(nc_sig), tax_ctrl$FeatureID)]
names(species_vec) <- rownames(nc_sig)

# Remove features with no species name
keep_rows <- which(!is.na(species_vec) & species_vec != "")
nc_sig <- nc_sig[keep_rows, , drop = FALSE]
species_vec <- species_vec[keep_rows]

# Aggregate FeatureIDs to Species
mat_species <- rowsum(nc_sig, group = species_vec)

# Sample metadata and type renaming
md <- as.data.frame(sample_data(ps_ctrl_abx2))

md$TypeRenamed <- md$Type
md$TypeRenamed <- gsub("^CTRL", "WT", md$TypeRenamed)
md$TypeRenamed <- gsub("ABX", "GMA", md$TypeRenamed)

keep_types <- c("WT_T1", "WT_T2", "WT_GMA_T1", "WT_GMA_T2")
keep_samples <- rownames(md)[md$TypeRenamed %in% keep_types]
keep_samples <- intersect(keep_samples, colnames(mat_species))

mat_species <- mat_species[, keep_samples, drop = FALSE]
md <- md[keep_samples, , drop = FALSE]

# Average across TypeRenamed
mat_long <- as.data.frame(t(mat_species))
mat_long$TypeRenamed <- md$TypeRenamed

mat_avg <- mat_long %>%
  dplyr::group_by(TypeRenamed) %>%
  dplyr::summarise(dplyr::across(where(is.numeric), mean)) %>%
  as.data.frame()

rownames(mat_avg) <- mat_avg$TypeRenamed
mat_avg <- mat_avg[, -1, drop = FALSE]

# Species as rows, Types as columns
mat_avg <- t(mat_avg)

desired_order <- c("WT_T1", "WT_T2", "WT_GMA_T1", "WT_GMA_T2")
desired_order <- desired_order[desired_order %in% colnames(mat_avg)]
mat_avg <- mat_avg[, desired_order, drop = FALSE]

# Log transform and row z-score
mat_log <- log10(mat_avg + 1)
mat_z <- t(scale(t(mat_log)))

# Row annotation: Phylum
species_to_phylum <- tax_ctrl %>%
  dplyr::filter(Species %in% rownames(mat_z)) %>%
  dplyr::select(Species, Phylum) %>%
  dplyr::distinct()

ann_row <- data.frame(Phylum = species_to_phylum$Phylum)
rownames(ann_row) <- species_to_phylum$Species
ann_row <- ann_row[rownames(mat_z), , drop = FALSE]

phy_lvls <- sort(unique(stats::na.omit(ann_row$Phylum)))
phy_cols_hm <- phylum_cols[phy_lvls]

# Column annotation
ann_col <- data.frame(
  Type = colnames(mat_z),
  Group = ifelse(colnames(mat_z) == "WT_GMA_T2", "WT_GMA", "WT")
)
rownames(ann_col) <- colnames(mat_z)

ann_colors <- list(
  Phylum = phy_cols_hm,
  Group = c(WT = "#999999", WT_GMA = "#000000")
)

# Plot heatmap
pal <- colorRampPalette(c("#2166AC", "white", "#B2182B"))

pheatmap(
  mat_z,
  color = pal(101),
  breaks = seq(-3, 3, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_row = ann_row,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 8,
  fontsize_col = 15,
  angle_col = 45,
  border_color = NA,
  main = "Significant Species (WT GMA vs WT, averaged by Type)"
)

# Save heatmap to Figures folder
ht <- grid::grid.grab()

grDevices::pdf(
  file = file.path(fig_dir, "WT_GMA_vs_WT_species_heatmap.pdf"),
  width = 10,
  height = 12,
  useDingbats = FALSE
)
grid::grid.draw(ht)
dev.off()

grDevices::tiff(
  file = file.path(fig_dir, "WT_GMA_vs_WT_species_heatmap.tiff"),
  width = 10,
  height = 12,
  units = "in",
  res = 300,
  compression = "lzw"
)
grid::grid.draw(ht)
dev.off()
```


```{r Heatmap_sig_species_FOP_GMA_vs_FOP_pool_AVERAGED, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Significant Feature IDs
sig_ids <- res_fop_abx2_merged %>%
  dplyr::filter(!is.na(padj), padj < 0.05) %>%
  dplyr::pull(FeatureID) %>%
  unique()

if (length(sig_ids) == 0) {
  stop("No significant features for FOP GMA vs FOP pool.")
}

# Normalized counts from DESeq object
nc <- counts(dds_fop_abx2, normalized = TRUE)
sig_ids <- intersect(sig_ids, rownames(nc))
nc_sig <- nc[sig_ids, , drop = FALSE]

# Map FeatureID to Species
tax_fop <- as.data.frame(tax_table(ps_fop_abx2))
tax_fop$FeatureID <- rownames(tax_fop)

species_vec <- tax_fop$Species[match(rownames(nc_sig), tax_fop$FeatureID)]
names(species_vec) <- rownames(nc_sig)

# Remove rows with no species name
keep_rows <- which(!is.na(species_vec) & species_vec != "")
nc_sig <- nc_sig[keep_rows, , drop = FALSE]
species_vec <- species_vec[keep_rows]

# Aggregate FeatureIDs to Species
mat_species <- rowsum(nc_sig, group = species_vec)

# Sample metadata and type renaming
md <- as.data.frame(sample_data(ps_fop_abx2))

md$TypeRenamed <- md$Type
md$TypeRenamed <- gsub("ABX", "GMA", md$TypeRenamed)

keep_types <- c("FOP_T1", "FOP_T2", "FOP_GMA_T1", "FOP_GMA_T2")
keep_samples <- rownames(md)[md$TypeRenamed %in% keep_types]
keep_samples <- intersect(keep_samples, colnames(mat_species))

mat_species <- mat_species[, keep_samples, drop = FALSE]
md <- md[keep_samples, , drop = FALSE]

# Average across TypeRenamed
mat_long <- as.data.frame(t(mat_species))
mat_long$TypeRenamed <- md$TypeRenamed

mat_avg <- mat_long %>%
  dplyr::group_by(TypeRenamed) %>%
  dplyr::summarise(dplyr::across(where(is.numeric), mean)) %>%
  as.data.frame()

rownames(mat_avg) <- mat_avg$TypeRenamed
mat_avg <- mat_avg[, -1, drop = FALSE]

# Species as rows, Types as columns
mat_avg <- t(mat_avg)

desired_order <- c("FOP_T1", "FOP_T2", "FOP_GMA_T1", "FOP_GMA_T2")
desired_order <- desired_order[desired_order %in% colnames(mat_avg)]
mat_avg <- mat_avg[, desired_order, drop = FALSE]

# Log transform and row z-score
mat_log <- log10(mat_avg + 1)
mat_z <- t(scale(t(mat_log)))

# Row annotation: Phylum
species_to_phylum <- tax_fop %>%
  dplyr::filter(Species %in% rownames(mat_z)) %>%
  dplyr::select(Species, Phylum) %>%
  dplyr::distinct()

ann_row <- data.frame(Phylum = species_to_phylum$Phylum)
rownames(ann_row) <- species_to_phylum$Species
ann_row <- ann_row[rownames(mat_z), , drop = FALSE]

phy_lvls <- sort(unique(stats::na.omit(ann_row$Phylum)))
phy_cols_hm <- phylum_cols[phy_lvls]

# Column annotation
ann_col <- data.frame(
  Type = colnames(mat_z),
  Group = ifelse(colnames(mat_z) == "FOP_GMA_T2", "FOP_GMA", "FOP")
)
rownames(ann_col) <- colnames(mat_z)

ann_colors <- list(
  Phylum = phy_cols_hm,
  Group = c(FOP = "#999999", FOP_GMA = "#000000")
)

# Plot heatmap
pal <- colorRampPalette(c("#2166AC", "white", "#B2182B"))

pheatmap(
  mat_z,
  color = pal(101),
  breaks = seq(-3, 3, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_row = ann_row,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 8,
  fontsize_col = 15,
  angle_col = 45,
  border_color = NA,
  main = "Significant Species (FOP GMA vs FOP, averaged by Type)"
)

# Save heatmap to Figures folder
ht <- grid::grid.grab()

grDevices::pdf(
  file = file.path(fig_dir, "FOP_GMA_vs_FOP_species_heatmap.pdf"),
  width = 10,
  height = 12,
  useDingbats = FALSE
)
grid::grid.draw(ht)
dev.off()

grDevices::tiff(
  file = file.path(fig_dir, "FOP_GMA_vs_FOP_species_heatmap.tiff"),
  width = 10,
  height = 12,
  units = "in",
  res = 300,
  compression = "lzw"
)
grid::grid.draw(ht)
dev.off()
```
```{r Heatmap_sig_species_FOP_vs_WT_pool_AVERAGED, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Significant Feature IDs
sig_ids <- phy_deseq_1_res_cft1 %>%
  dplyr::filter(!is.na(padj), padj < 0.05) %>%
  dplyr::pull(FeatureID) %>%
  unique()

if (length(sig_ids) == 0) {
  stop("No significant features for FOP vs WT pool.")
}

# Normalized counts from DESeq object
nc <- counts(phy_deseq_1, normalized = TRUE)
sig_ids <- intersect(sig_ids, rownames(nc))
nc_sig <- nc[sig_ids, , drop = FALSE]

# Map FeatureID to Species
tax_all <- as.data.frame(tax_table(ps_filtered))
tax_all$FeatureID <- rownames(tax_all)

species_vec <- tax_all$Species[match(rownames(nc_sig), tax_all$FeatureID)]
names(species_vec) <- rownames(nc_sig)

keep_rows <- which(!is.na(species_vec) & species_vec != "")
nc_sig <- nc_sig[keep_rows, , drop = FALSE]
species_vec <- species_vec[keep_rows]

# Aggregate FeatureIDs to Species
mat_species <- rowsum(nc_sig, group = species_vec)

# Sample metadata and type renaming
md <- as.data.frame(sample_data(ps_filtered))

md$TypeRenamed <- md$Type
md$TypeRenamed <- gsub("^CTRL", "WT", md$TypeRenamed)
md$TypeRenamed <- gsub("ABX", "GMA", md$TypeRenamed)

keep_types <- c(
  "WT_T1", "WT_T2", "WT_GMA_T1",
  "FOP_T1", "FOP_T2", "FOP_GMA_T1"
)

keep_samples <- rownames(md)[md$TypeRenamed %in% keep_types]
keep_samples <- intersect(keep_samples, colnames(mat_species))

mat_species <- mat_species[, keep_samples, drop = FALSE]
md <- md[keep_samples, , drop = FALSE]

# Average across TypeRenamed
mat_long <- as.data.frame(t(mat_species))
mat_long$TypeRenamed <- md$TypeRenamed

mat_avg <- mat_long %>%
  dplyr::group_by(TypeRenamed) %>%
  dplyr::summarise(dplyr::across(where(is.numeric), mean)) %>%
  as.data.frame()

rownames(mat_avg) <- mat_avg$TypeRenamed
mat_avg <- mat_avg[, -1, drop = FALSE]

# Species as rows, Types as columns
mat_avg <- t(mat_avg)

desired_order <- c(
  "WT_T1", "WT_T2", "WT_GMA_T1",
  "FOP_T1", "FOP_T2", "FOP_GMA_T1"
)
desired_order <- desired_order[desired_order %in% colnames(mat_avg)]
mat_avg <- mat_avg[, desired_order, drop = FALSE]

# Log transform and row z-score
mat_log <- log10(mat_avg + 1)
mat_z <- t(scale(t(mat_log)))

# Row annotation: Phylum
species_to_phylum <- tax_all %>%
  dplyr::filter(Species %in% rownames(mat_z)) %>%
  dplyr::select(Species, Phylum) %>%
  dplyr::distinct()

ann_row <- data.frame(Phylum = species_to_phylum$Phylum)
rownames(ann_row) <- species_to_phylum$Species
ann_row <- ann_row[rownames(mat_z), , drop = FALSE]

phy_lvls <- sort(unique(stats::na.omit(ann_row$Phylum)))
phy_cols_hm <- phylum_cols[phy_lvls]

# Column annotation
ann_col <- data.frame(
  Type = colnames(mat_z),
  Group = ifelse(grepl("^WT", colnames(mat_z)), "WT", "FOP")
)
rownames(ann_col) <- colnames(mat_z)

ann_colors <- list(
  Phylum = phy_cols_hm,
  Group = c(WT = "#999999", FOP = "#000000")
)

# Plot heatmap
pal <- colorRampPalette(c("#2166AC", "white", "#B2182B"))

pheatmap(
  mat_z,
  color = pal(101),
  breaks = seq(-3, 3, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  annotation_row = ann_row,
  annotation_col = ann_col,
  annotation_colors = ann_colors,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 8,
  fontsize_col = 15,
  angle_col = 45,
  border_color = NA,
  main = "Significant Species (FOP vs WT pool, averaged by Type)"
)

# Save heatmap to Figures folder
ht <- grid::grid.grab()

grDevices::pdf(
  file = file.path(fig_dir, "FOP_vs_WT_pool_species_heatmap.pdf"),
  width = 10,
  height = 12,
  useDingbats = FALSE
)
grid::grid.draw(ht)
dev.off()

grDevices::tiff(
  file = file.path(fig_dir, "FOP_vs_WT_pool_species_heatmap.tiff"),
  width = 10,
  height = 12,
  units = "in",
  res = 300,
  compression = "lzw"
)
grid::grid.draw(ht)
dev.off()
```